---
title: "Week 2 Lecture Notes"
author: "Dustin Duncan"
format: 
  html:
    code-fold: true
    toc: true
    number-sections: true
    embed-resources: true
editor: visual
execute:
  echo: true
  message: false
  warning: false
---

```{r}
library(tidyverse)
# install.packages("nloptr")
library(nloptr)
library(knitr)
```

## Examples of Natural Resource Economics in R

-   We get the specific functional form and the parameters that we're using, then we build the model in R with a custom function (I think)

### Example 1: optimal forest rotation in R

We seek to select the optimal harvest age T\* for an even-aged forest to maximize the present discounted value of profits from a single rotation. The timber yield is given by $Q(T)=e^{a-b/T}$, where $a=10, b=53.27$, the cost of planting the stand is $c=300$, the price of a unit of timber is $p=1$, and the discount rate is $\delta=0.05$. Thus, we choose $T^*$ to maximize $PVNB(T^*)=pQ(T^*)e^{-\delta T^*} - c$ where $T^* \gt 0$.

### A.

What is the optimal rotation $T^*$?

[**Solution**]{style="color: blue;"}

[Build optimization problem with an objective function and constraint ensuring that harvest age $T^*$ is strictly positive]{style="color: blue;"}

```{r}

# A code chunk starts with '''{r} and ends with '''.  You can run a code chunk by clicking the arrow in the upper right (after you've run the libraries code chunk)

# The statement below designates a function named obj_fcn_tree to depend on the choice variable, in this case the rotation length, and a set of parameters.  "choice" needs to be in the first position, but it doesn't have to have this name.  Note the use of <- rather than =

obj_fcn_tree <- function(choice,a,b,c,p,discount) { 
  
  Q=exp(a-b/choice) # 'exp' stands in for 'e^X'
  benefits = p*Q*exp(-discount*choice)-c # Discount is the delta from above 
  ## Benefits are the present value of net benefits (PVNB in the above formula)
  
  ## Notice the minus sign in front of benefits.  This is added because we want nloptr to find the maximum of the function; it always find the minimum.

  return(-benefits)
}


## Optimization ##

# Always include this options statement
# "NLOPT_LN_COBYLA"
options=list("algorithm"="NLOPT_LN_COBYLA", # Options are arguments for nloptr
             # COBYLA: "Constrained Optimization by Linear Approximation"
             xtol_rel=1e-8, # stopping criterion for relative change reached (default is 1e-4)
             maxeval=16000 # maximum number of function evaluations 
             )

# This next statement labels the output from running nloptr as out_1a.  Before running it the first type, load the package by typing install.packages("nloptr") in the command line below.  
# Inside the parentheses we define 
## 1) the initial value of choice, 
## 2) the function we want evaluated, 
## 3) the lower bound on choice; can also specify the upper bound with ub=, 
## 4) the options list, 
## 5) the values of the parameters for the function. Type ??nloptr in the command line to get to a help page.

out_1a=nloptr(x0=1, # Vector with starting values for optimization
          eval_f=obj_fcn_tree, # Function that returns the value of the objective function (benefits)
          lb=0, # Vector with lower bounds of the controls
          opts = options, # 'algorithm' is required, these control the termination conditions 
          discount=0.05,a=10,b=53.27,c=300,p=1 # These are the arguments that are provided to the objective function 
          )


# There are lots of values you can print out after running nloptr.  The statement below prints out the value of choice that maximizes benefits.  The command refers to the output label "out_1a" and what should be printed "solution", separated by a $,  In the Global Environment, click on the search icon to the right of out_1a after you run the code chunk and you can see what else can be printed out.  

cat("The optimal harvest age for timber in an even-aged forest is:", out_1a$solution, "\n")

cat("The value of the objective function (PVNB) at the optimum year is:", -out_1a$objective, "\n")
out_1a$message


```

## A quick explanation of loops

We will often solve problems that involving summations over time. As a simple example, how would you sum the numbers between 1 and 100?

```{r}

total=0

for(i in 1:100){

total=total+i  
  
  
}

print(total)

```

## Example 2: Simple fishery

This problem involves a fish stock that evolves according to $X_{t+1}=X_t+F(X_t)-Y_t$, where the growth function is given by $F(X_t)=rX_t(1- \frac{X_t}{K})$. The profits from the fishery in each period are $pY_t-(\frac{c}{2})* \frac{Y_t^2}{X_t}$ and the catch is proportional to the stock: $Y_t=\alpha X_t$. The objective is to maximize the present value of net benefits (PVNB) from time 0 to time 19 by selecting the catchability coefficient $\alpha$:

$$
\max_{\alpha} \sum^{t=19}_{t=0}\rho^t(pY_t-(\frac{c}{2})*\frac{Y_t^2}{X_t})
$$

subject to $X_{t+1}=X_t+rX_t(1-X_t/K)-Y_t$ and $X_0$ given. Use the following parameter values: $\rho=\frac{1}{1.05}$, p=5, c= 1, r=1, K=1, $X_0=0.1$.

### A.

What is the optimal value of the catchability coefficient?

[**Solution** ]{style="color: blue;"}

```{r}


# For this problem, the function is obj_fcn_fish, which depends on the choice variable (choice) and parameters, which includes the initial state of the system, or X_0, and the length of the planning horizon (19), labeled "period". 

obj_fcn_fish <- function(choice,int_state,discount,p,c,r,K,period) {
  
# For this problem, we have 20 values for each X (state), Y (yield), and \pi (benefits), and so we need to designate variables as vectors.  mode says the vector will be populated with numbers, a length of zero lets us define the length (20) later  
  
  state<-vector(mode="numeric", # Mode says vector will be populated with numbers
                length=0) # Setting length to zero lets us define it later 
  yield<-vector(mode="numeric",length=0)
  benefits<-vector(mode="numeric",length=0)
  
# The statements below define the three variables in the initial period, which must be designated as a 1 even though our sum starts at 0.  
  
  
  state[1]=int_state
  yield[1]=choice*state[1]
  benefits[1]=p*yield[1]-(c/2)*(yield[1]^2)/state[1]
  
# To define the three variables in the remaining periods, I'm using a loop from 2 to period+1 (below, I will define period to equal 19).  Notice that the loop starts and ends with braces.  We're still within the loop defining the function.
  
  for(i in 2:(period+1)){ # Remember period is the length of the planning horizon
    
# The following three equations define the state, yield, and benefit in a given period i.  Notice that the state equation for period i depends on values in period i-1.
    
    state[i]=state[i-1]+r*state[i-1]*(1-state[i-1]/K)-yield[i-1] 
    # The state in period i depends on the previous period's (i-1) state
    yield[i]=choice*state[i] # Yield depends on the choice and the state in period i
    benefits[i]=p*yield[i]-(c/2)*(yield[i]^2)/state[i] 
    # Benefits depend on the yield in i, and the state in i, with the parameters p and     c included 
       
             
  }
  
  
# The following statement defines a sequence of integers from 0 to period (i.e., 0, 1, ...).  Subsequent statements discount benefits, add those values up using the sum function, and then returns the object to be minimized by nloptr: -npv
  
  t=seq(from=0,period) # Defines the sequence of integers in period 
  
  pv=discount^t*benefits # This is the formula for discounting benefits over time
    
  npv=sum(pv) # This adds the discounted benefits up and is the goal of our objective function, to maximize this sum by choosing the optimal 'alpha' or harvest rate 
  
  return(-npv)
}


# I will always give you the options list

local_opts<-list("algorithm"="NLOPT_LN_COBYLA",xtol_rel=1e-15)
options=list("algorithm"="NLOPT_LN_AUGLAG",xtol_rel=1e-15,maxeval=16000,"local_opts"=local_opts)





out_2a=nloptr(x0=0.5, # Starting values for the optimization
              eval_f=obj_fcn_fish, # Function that we defined above. 
              ## "Calculate the state of the fishery over time, and the benefits for a                   choice over time, by choosing an optimal catchability coefficient,                    alpha" 
              lb=0, # Lower bound is zero 
              opts = options, # Options defined above 
              int_state=0.1,discount=1/1.05,p=5,c=1,r=1,K=1,period=19) # Function arguments 


cat("The optimal alpha term to maximize benefits over time in this fishery is:", out_2a$solution )
cat("By choosing the optimal alpha term, the total net present benefits will be:", -out_2a$objective )



```

```{r}
## Function for making a table; I'm designating the values of parameters here.

stock<-function(choice,int_state=0.1,p=5,c=1,r=1,K=1,discount=1/1.05,period=19){
  
  
    #make storage vectors
  state<-vector(mode="numeric",length=0)
  yield<-vector(mode="numeric",length=0)
  benefits<-vector(mode="numeric",length=0)
  

  #initial state and benefits
  
  
  state[1]=int_state
  yield[1]=choice*state[1]
  benefits[1]=p*yield[1]-(c/2)*(yield[1]^2)/state[1]
  
  for(i in 2:(period+1)){
    
# Add the stock dyanmics
    
  state[i]=state[i-1]+r*state[i-1]*(1-state[i-1]/K)-yield[i-1]
  yield[i]=choice*state[i]
  benefits[i]=p*yield[i]-(c/2)*(yield[i]^2)/state[i]
             
  }
  
  #define period and discount
  t=seq(from=0,period)
  
  pv=discount^t*benefits
  
  npv=sum(pv)

# This time we return a data frame with the variables we want in the table.  The value after the = indicates the column label.
  
  return(data.frame(t=t,PV=pv,X=state,Y=yield))

}

# The next command is telling R to use the solution found above, choice, to evaluate the function stock and to round the values in the data frame to two places past the decimal point.  The second command says to make a table with these values. 

tbl_2a=round(stock(choice=out_2a$solution),2)      #  2 indicates rounding units

kable(tbl_2a)


```

### B.

What is the optimal catchability coefficient $\alpha$ when different values can be selected in each period?

[**Solution** ]{style="color: blue;"}

```{r less_simple}

obj_fcn_fish<-function(choice,int_state,p,c,r,K,discount,period){
  
  
    #make storage vectors
  state<-vector(mode="numeric",length=0)
  yield<-vector(mode="numeric",length=0)
  benefits<-vector(mode="numeric",length=0)
  
  

  #initial state and benefits; notice that choice is now indexed by 1 (and i below) because there will be 20 different values.
  
  
  state[1]=int_state
  yield[1]=choice[1]*state[1]
  benefits[1]=p*yield[1]-(c/2)*(yield[1]^2)/state[1]
  
  for(i in 2:(period+1)){
    
# Add the stock dynamics
    
    state[i]=state[i-1]+r*state[i-1]*(1-state[i-1]/K)-yield[i-1]
    yield[i]=choice[i]*state[i]
    benefits[i]=p*yield[i]-(c/2)*(yield[i]^2)/state[i]
       
             
  }
  
  
  #define period
  t=seq(from=0,period)
  
  pv=discount^t*benefits
    
  npv=sum(pv)
  
  return(-npv)
  
  
}


local_opts<-list("algorithm"="NLOPT_LN_COBYLA",xtol_rel=1e-15)
options=list("algorithm"="NLOPT_LN_AUGLAG",xtol_rel=1e-15,maxeval=16000,"local_opts"=local_opts)


# Notice that below you need to specify twenty initial value of choice, which you can do with the rep statement; same goes for lower and upper bounds.


out_2b=nloptr(x0=rep(0.5,times=20), # x0 is the starting value for choice, and replicating it 20 times makes it a vector 
              eval_f=obj_fcn_fish,
              lb=rep(0,times=20),
              ub=rep(1,times=20),
              opts = options,int_state=0.1,discount=1/1.05,p=5,c=1,r=1,K=1,period=19)


print ( -out_2b$objective )
print ( out_2b$solution )

```

```{r}
## Function for making a table

stock<-function(choice,int_state=0.1,p=5,c=1,r=1,K=1,discount=1/1.05,period=19){
  
  
    #make storage vectors
  state<-vector(mode="numeric",length=0)
  yield<-vector(mode="numeric",length=0)
  benefits<-vector(mode="numeric",length=0)
  

  #initial state and benefits; notice that choice is now indexed by time
  
  
  state[1]=int_state
  yield[1]=choice[1]*state[1]
  benefits[1]=p*yield[1]-(c/2)*(yield[1]^2)/state[1]
  
  for(i in 2:(period+1)){
    
# Add the stock dyanmics
    
  state[i]=state[i-1]+r*state[i-1]*(1-state[i-1]/K)-yield[i-1]
  yield[i]=choice[i]*state[i]
  benefits[i]=p*yield[i]-(c/2)*(yield[i]^2)/state[i]
             
  }
  
  #define period and discount
  t=seq(from=0,period)
  
  pv=discount^t*benefits
  
  
  return(data.frame(t=t,PV=pv,X=state,Y=yield,alpha=choice))

}

tbl_2b=round(stock(out_2b$solution),1)

kable(tbl_2b)
```

## Example 3: Stock Pollutant {.tabset}

This problem involves a stock pollutant that evolves according to $Z_{t+1}=z_t(1-\gamma)+\alpha*q_t$, where $\gamma$ captures how much pollution degrades each period and $\alpha$ captures how much pollution is produced by production $q_t$. Suppose that production is worth $p$ per unit of output, so net revenues are $pq$. Damages caused by pollution are $C=aZ_t - bZ_t^2$. The objective is to maximize the present value of net benefits (PNVB) from time 0 to time 10:

$$
\mathrm{\max_q \sum^{t=10}_{t=0}\rho^t(pq_t-(aZ_t-bZ_t^2))}
$$

subject to $Z_{t+1}=z_t(1-\gamma)+\alpha*q_t$ and $Z_0$ given. Use the following parameter values: $\rho=\frac{1}{1.05}$, p=100, a= 2, b=-0.25,$\gamma=0.2$, $\alpha=2$, and $Z_o$=20.

### A.

What is the amount of production $q_t$ in each period according to the solution?

[**Solution** ]{style="color: blue;"}

```{r}


# Notice that period is not included as a parameter of the function

obj_fcn_stock<-function(choice,int_state,a,b,gamma,p,alpha,discount){

  
# The state and benefits vectors will have the same length as the choice vector    
  
    #make storage vectors
  state<-vector(mode="numeric",length=length(choice))
  benefits<-vector(mode="numeric",length=length(choice))
  

  #initial state and benefits
  
  
  state[1]=int_state
  
  benefits[1]=p*choice[1]-(a*state[1]-b*state[1]^2)

#  The loop goes from 2 to whatver the length of choice ends up being
    
  for(i in 2:length(choice)){
    
# Add the stock dyanmics
    
    state[i]=state[i-1]*(1-gamma)+alpha*choice[i-1] #Add the time from the choice to the state
    
    
  benefits[i]=p*choice[i]-(a*state[i]-b*state[i]^2)
       
             
  }
  
  
  #define period
  
# Populating a vector with 0,1,... that ends with the length of the benefits vector
  
  t=seq(from=0,length.out=length(benefits))
  
  pv=discount^t*benefits
    
  npv=sum(pv)
  
  return(-npv)
  
  
}


local_opts<-list("algorithm"="NLOPT_LN_COBYLA",xtol_rel=1e-15)
options=list("algorithm"="NLOPT_LN_AUGLAG",xtol_rel=1e-15,maxeval=16000,"local_opts"=local_opts)

# Because I didn't include period as a parameter in the function, I need to define it here.

period=11

out_3a=nloptr(x0=rep(1,period),eval_f=obj_fcn_stock,
              lb=rep(0,period), # Lower bound, cannot have negative extraction
              ub=rep(10,period), # Upper bound, extraction cannot exceed 10
              opts = options,
              int_state=20,
              discount=1/1.05,
              a=2,
              b=-.25,
              gamma=0.2,
              alpha=2,
              p=100)


ta_solution=round(out_3a$solution,5)

ta_value=round(-out_3a$objective,3)

```

[ Comments can be added here ]{style="color: blue;"}

```{r}
## Function for making a table

stock<-function(choice,int_state=20,a=2,b=-0.25,gamma=.2,p=100,alpha=2,discount=1/1.05){
  
  
    #make storage vectors
  state<-vector(mode="numeric",length=length(choice))
  benefits<-vector(mode="numeric",length=length(choice))
  

  #initial state and benefits
  
  
  state[1]=int_state
  
  benefits[1]=p*choice[1]-(a*state[1]-b*state[1]^2)
  
  for(i in 2:length(choice)){
    
# Add the stock dyanmics
    
    state[i]=state[i-1]*(1-gamma)+alpha*choice[i-1] #Add the time from the choice to the state
    
  benefits[i]=p*choice[i]-(a*state[i]-b*state[i]^2)
             
  }
  
  #define period and discount
  t=seq(from=0,length.out=length(benefits))
  
  pv=discount^t*benefits
  
  
  return(data.frame(t=t,PV=pv,Z=state,q=choice))

}

tbl_3a=round(stock(out_3a$solution),1)

kable(tbl_3a)
```
