---
title: "EM 242 HW 3"
author: "Andrew Plantinga"
date: "2024-10-22"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
library(nloptr)
library(knitr)
```

## Question 1: Forestry Rotation {.tabset}

This problem expands on the forest rotation problem presented in class.  Suppose that a stand of trees is planted in time 0 at a cost of c, grown for $T_1$ years, harvested, replanted at cost c, grown for $T_2$ years, harvested, replanted, and so on.  If we complete six rotations, what should be the values of $T_1$, $T_2$, $T_3$, $T_4$, $T_5$, and $T_6$?  Assume that different values can be chosen for each year and that the stand does not need to be replanted after the sixth harvest.  Use the parameter values a=10, b=53.27, c=250, p=1.5, and $\delta$=0.05.

### A.

Before you solve the problem for six rotations, find the optimal rotation when only one rotation is done.  What is the optimal value of $T_1$?

<span style="color: blue;">**The optimal rotation length when only one rotation is done is 32.64**
</span>

$$
\mathrm{ Q(T)=e^{a-b/T} \\ 
\max_{T_1} ~\Pi_{1}=pQ(T_{1})e^{-\delta T_{1}}-c}
$$ 

```{r}

forest_fct <- function(choice, c, p, a, b, discount) {
  
  Q=exp(a-b/choice)
  benefits=p*Q*exp(-discount*choice) - c
  return(-benefits)
}

options=list("algorithm"="NLOPT_LN_COBYLA",xtol_rel=1e-8,maxeval=16000)

# Use the parameter values a=10, b=53.27, c=250, p=1.5, and $\delta$=0.05
out_1a=nloptr(x0=1,
              eval_f=forest_fct,
              opts=options,
              lb=0,
              a=10,
              b=53.27,
              c=250,
              p=1.5,
              discount=.05
              )

out_1a$solution
-out_1a$objective

```


### B. 

Now solve the problem for six consecutive rotations.  What are the six rotation lengths that you found?

<span style="color: blue;">**The six rotation lengths that were found when 6 rotations are done are: 29.32, 29.32, 29.35, 29.48, 30.04, and 32.65.**
</span>


```{r}

forest_b <- function(choice, a, b, c, p, discount, period) {
  
  ## initiating vectors for state, benefits, and quantity harvested 
  state=vector(mode="numeric", length=0)
  benefits=vector(mode="numeric", length=0)
  Q=vector(mode="numeric", length=0)
  
  # state[1]=choice[1] # Choice 1 is the first state. Length of first rotation
  Q[1]=exp(a-b/choice[1]) # Quantity harvested at t1 depends on length of first rotation
  benefits[1]=p*Q[1]*exp(-discount*choice[1])-c # benefits discounted to choice of length of time for 1st rotation 
  state[1]=choice[1]
    
  for(i in 2:period){
    
    state[i]=state[i-1] + choice[i] # State in this case is the total time passed
    
    Q[i]=exp(a-b/choice[i]) # Quantity harvested depends on this time choice
    # option for c: c*exp(-discount*state[i-1]))
    benefits[i]=p*Q[i]*exp(-discount*state[i]) - c*exp(-discount*state[i-1])
    # /(1-exp(-discount*state[i])) 
    # Benefits are discounted to current total time passed (state[i]), and costs are discounted to the time passed when costs were incurred, which is state[i-1]
  }
  
  npv=sum(benefits)
  
  return(-npv)
}


options=list("algorithm"="NLOPT_LN_COBYLA",xtol_rel=1e-8,maxeval=16000)

period=6

# Use the parameter values a=10, b=53.27, c=250, p=1.5, and $\delta$=0.05
out_1b=nloptr(x0=rep(20, period),
              eval_f=forest_b,
              opts=options,
              lb=rep(0, period),
              a=10,
              b=53.27,
              c=250,
              p=1.5,
              period=6,
              discount=0.05
              )

cat("The optimal rotation lengths when six rotations are done are:\n ",round(out_1b$solution, 2), "\n")
cat("The Present value of six consecutive rotations is:",round(-out_1b$objective, 2))
```



### C.

Explain why $T_1$ through $T_5$ are shorter than $T_6$?

<span style="color: blue;">**T_6 is the longest rotation length because there are no costs associated with replanting after the 6th rotation. As a result, the first five rotations are shorter due to costs associated with replanting, and the discounted benefits associated with their lengths; so shortening the rotation length compared to period six increases their net present benefits.**
</span>

<span style="color: blue;">

</span>

### D.

What happens to $T_1$ through $T_5$ when the cost of replanting the stand increases to 500? Explain. Why doesn't $T_6$ change with the higher replanting cost?


```{r}

options=list("algorithm"="NLOPT_LN_COBYLA",xtol_rel=1e-8,maxeval=16000)

out_1d=nloptr(x0=rep(20, period),
              eval_f=forest_b,
              opts=options,
              lb=rep(0, period),
              a=10,
              b=53.27,
              c=500,
              p=1.5,
              period=6,
              discount=0.05
              )
cat("The optimal rotation lengths for six rotations when the cost of replanting \n is $500 are:", round(out_1d$solution, 2), "\n Compared to when the cost was $250:", round(out_1b$solution, 2), "\n")
cat("The net present value of benefits for six rotations when the cost of replanting \n is $500 is:", round(-out_1d$objective, 2))

```

<span style="color: blue;">**When the cost of replanting is increased to 500, increasing the rotation length allows for the additional benefits from the timber harvest to counteract the increased costs of replanting. Because there are still no costs associated with replanting at the end of the sixth rotation, the increased  cost would have no effect on its optimal rotation length.**
</span>

<span style="color: blue;">


</span>



### E.

Compare your answer to part a (the single rotation) to $T_6$. Why are they the same?

<span style="color: blue;">**The optimal rotation length for a single rotation is the same as the sixth rotation length when there are six rotations done. This is due to the costs of replanting not playing a factor in either scenario and both scenario's rotation lengths being optimized independently, so discounting of benefits of the harvest leads to both optimal rotation lengths being 32.65 years.**
</span>

### F

Now solve for the optimal rotation when an infinite number of rotations are done.  Compute the present value of net revenues (i.e., $objective). Compare this to the present value of net revenues when only six rotations are done. Are they close in magnitude?  Why?

<span style="color: blue;">**The present value of net revenues when an infinite number of rotations are done is within 1 unit difference from the present value of net revenues when 6 rotations are done due to the discounting factor. The two values being similar implies that there is very small value added after the sixth rotation in the infinite rotation scenario due to the discount factor's effect over longer time periods.**</span>

```{r}
inft_fct = function(choice, a, b, c, p, discount) {
  
  Q=exp(a-b/choice)
  benefits=(p*Q*exp(-discount*choice)-c)/(1-exp(-discount*choice))
  
  return(-benefits)
}
options=list("algorithm"="NLOPT_LN_COBYLA",xtol_rel=1e-8,maxeval=16000)

out_1f=nloptr(x0=20,
              eval_f=inft_fct,
              opts=options,
              lb=0,
              a=10,
              b=53.27,
              c=250,
              p=1.5,
              discount=0.05
              )
cat("The optimal rotation length for infinite rotations is:",round(out_1f$solution, 2), "\n")
cat("The net present value of six rotations was: ", round(-out_1b$objective, 2), "\n", "and the net present value of infinite rotations is:", round(-out_1f$objective, 2))
```

## Question 2: Varying Initial Age {.tabset}

In this problem, you are asked to find the optimal rotation when the initial age of the stand is not zero.  The key question is whether a positive initial age should change the solution.  That is, if T* is the optimal rotation age for a stand starting at age 0, will it still be the same optimal rotation age for a stand starting at age $A$>0?  The volume of timber evolves according to $Q(T)=e^{a-b/T}$  where a=13, b=185, and T is the age of the trees. The price of timber is p=1.78 and the cost of planting the stand at the start of each rotation is 1000.  The discount rate is $\delta$=0.05.  

### A.
	
If the stand is grown for an infinite number of rotations, what is the optimal rotation length when the initial age of the stand is zero?  Call this value $T^*$.  What is the present discounted value of net timber revenues from an infinite number of rotations?  Call this value $\pi_\infty$.

<span style="color: blue;">**The optimal rotation length for infinite rotations when the initial age of the stand is zero is 60.14, and the net present value of benefits is 837.67**</span>



```{r}

inft_fct2 <- function(choice, int_age, a, b, c, p, discount) {
  
  Q=exp(a-b/(choice+int_age))
  benefits=(p*Q*exp(-discount*(choice))-c)/(1-exp(-discount*(choice)))
  
  return(-benefits)
}

options=list("algorithm"="NLOPT_LN_COBYLA",xtol_rel=1e-15,maxeval=16000)

out_2a=nloptr(x0=1, 
              eval_f=inft_fct2,
              opts=options,
              lb=0,
              a=13,
              b=185,
              c=1000,
              p=1.78,
              discount=0.05,
              int_age=0)
cat("The optimal rotation length for infinite rotations when the initial \n age of the stand is zero is: ",round(out_2a$solution, 2), "\n")
cat("Leading to the net present value of benefits being: ", round(-out_2a$objective, 2))

```



### B.

Suppose that the initial age of the stand is A and $T_1$ is the additional number of years the stand is grown until harvest. Then, $A+T_1$ is the age of the stand at the end of the first rotation. Assuming the first rotation is followed by an infinite number of rotations, write an expression for present discounted value of net timber revenues from an infinite number of rotations (including the first one).

$$
\mathrm{\Pi_{\infty}= pQ(T_{1}+A)e^{-\delta T_{1}}-c+\Pi_{\infty}e^{-\delta T_{1}}}
$$


<span style="color: blue;">**See formula above**</span>

<span style="color: blue;">

</span>

### C.

Using your formula in b, and assuming A=30, what is the optimal value of $T_1$? How does your answer change when A=40?  A=60.14?  A=90?  Make a table showing your results.

<span style="color: blue;">**Assuming A=30, the optimal value of T_1 would be 30.91; when A=40, T_1 is 20.87; when A=60.14, T_1 is 0.69; when A=90, T_1 is 0.**</span>



```{r}

fct_2c <- function(choice, int_age, a, b, c, p, discount, pi_inf) {
  
  Q=exp(a-b/(choice+int_age))
  # b1=(p*Q*exp(-discount*(choice))-c)/(1-exp(-discount*(choice)))
  benefits=(p*Q*exp(-discount*choice)-c) + pi_inf*(1-exp(-discount*choice))
  
  return(-benefits)
}



out_2c = nloptr(x0=10,
                eval_f = fct_2c,
                opts=options,
                lb=0,
                int_age=30,
                a=13,
                b=185,
                c=1000,
                p=1.78,
                discount=0.05,
                pi_inf=round(-out_2a$objective, 2))
# out_2c$solution
# -out_2c$objective

out_2c2 = nloptr(x0=10,
                eval_f = fct_2c,
                opts=options,
                lb=0,
                int_age=40,
                a=13,
                b=185,
                c=1000,
                p=1.78,
                discount=0.05,
                pi_inf=round(-out_2a$objective, 2))
# out_2c2$solution
# -out_2c2$objective

out_2c3 = nloptr(x0=10,
                eval_f = fct_2c,
                opts=options,
                lb=0,
                int_age=60.14,
                a=13,
                b=185,
                c=1000,
                p=1.78,
                discount=0.05,
                pi_inf=round(-out_2a$objective, 2))
# out_2c3$solution
# -out_2c3$objective

out_2c4 = nloptr(x0=10,
                eval_f = fct_2c,
                opts=options,
                lb=0,
                int_age=90,
                a=13,
                b=185,
                c=1000,
                p=1.78,
                discount=0.05,
                pi_inf=round(-out_2a$objective, 2))
# out_2c4$solution
# -out_2c4$objective

table <- data.frame("A"=c(30, 40, 60.14, 90), "T1"=c(round(out_2c$solution, 2), round(out_2c2$solution, 2), round(out_2c3$solution,2), round(out_2c4$solution, 2)), "Objective"=c(round(-out_2c$objective, 2), round(-out_2c2$objective,2), round(-out_2c3$objective,2), round(-out_2c4$objective, 2)))

knitr::kable(table, align='c')
```

### D.

What do you notice about $A+T_1$?  Explain.

<span style="color: blue;">**When t_1 is calculated with A=0, it states that the optimal rotation length is 60.14. This is the optimal rotation length to maximize benefits from harvest when considering discounting. As a result, with increasing initial age A, the optimal solution maximizes the harvest in the same way, while decreasing the rotation length to decrease the net present value of benefits lost to discounting. For A=0 through A=60.14, the optimal solution states that the present value of net benefits for a rotation is reached when the total age of the stand is 60.14; as a result, when A is less than 60.14, the optimal rotation length will be (close to) the difference between 60.14 and A to minimize value lost from discounting. When the stand's initial age is A=90, the present value of net benefits is maximized by harvesting the stand immediately; the age of the stand is older than the optimal age, so it is harvested immediately, not discounted, and its PVNB is higher than the other scenarios. **</span>

<span style="color: blue;">


</span>


