---
title: "Homework 4"
author: "Andrew Plantinga"
date: "2024-10-29"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
library(nloptr)
library(knitr)
library(cowplot)
```

## Question 1: Costless Exhaustible Resource Extraction {.tabset} 

Suppose there is an exhaustible resource that is costlessly extracted.  The inverse demand curve for the resource is  $p_t=a-bq_t$ where $q_t$ is the quantity extracted in time t and a=40, b=1.  The discount rate is $\delta=0.05$  and the initial stock of reserves is $R_0$ = 778.1259 units.  Set up a program to find the competitive market equilibrium that maximizes the present discounted profits from resource extraction.  $R_t$ is the reserves remaining in year t,  $\pi_t$ is the profits in each year (without costs, they are just equal to $p_tq_t$ ), and $\lambda_{t+1}=\frac{p_{t}}{\rho}$. 


### A

List the values of $t$, $p_t$, $q_t$, $R_t$, $\lambda_t$, and  $\pi_t$ in a table below. What should be the final (year T) values of $p_T$ and $R_T$? Why?  

<span style="color: blue;">**The final values of  p_T and R_T should be 40 and 0 respectively. To maximize the present discounted profits from resource extraction in a competitive market, the firm would harvest all of the reserves available, leaving none left at the end. In addition, in competitive equilibrium for an exhaustible resource, the price of the good should increase at the rate of interest, or discount in this case until it reaches the choke point (40), where demand is zero.**</span>


```{r}
opt_fct <- function(choice, int_state, a, b, c, discount) {
  
  t=choice
  P=vector(mode="numeric", length=0) # P is price
  Q=vector(mode="numeric", length=0) # This is q
  R=vector(mode="numeric", length=0) # State is reserves
  benefits=vector(mode="numeric", length=0)
  # lambda=vector(mode="numeric", length=0)

  # Define initial conditions. Since we are working down these are at the end of the time frame
  P[t+1]=a
  R[t+1]=0
  Q[t+1]=0
  benefits[t+1]=0

  for(i in (t+1):2) {

    P[i-1]=(P[i]-(c*discount))/(1+discount) # Gives P from t+1 to 1

    Q[i-1]=(a-P[i-1])/b # Gives Q from t+1 to 1

  }
  
  benefits[1]=(P[1]-c)*Q[1]

  R[1]=int_state
  
  for(i in 2:((t+1)-1)) {
    R[i]=R[i-1]-Q[i-1] # Gives R from 2 to t because we know it in t+1

    benefits[i]=(P[i]-c)*Q[i] # Gives benefits from 2 to t for same as above
  }

  time=seq(from=0,to=t)
  rho=1/(1+discount)
  pv=rho^time*benefits
  npv=sum(pv)

  return(-npv)
}



constraint <- function(choice, int_state, a, b, c, discount) {
  t=choice
  time=seq(from=0,to=t)
  R=vector(mode="numeric",length=0)
  const=vector(mode="numeric",length=0)
  P=vector(mode="numeric",length=0)
  Q=vector(mode="numeric",length=0)
  
  P[t+1]=a
  R[t+1]=0
  Q[t+1]=0
  
  for(i in (t+1):2) {

    P[i-1]=(P[i]-(c*discount))/(1+discount) # Gives P from t+1 to 1

    Q[i-1]=(a-P[i-1])/b # Gives Q from t+1 to 1

  }
  
  # Defining the constraint 
  const=int_state-sum(Q)

  return(const)

}

local_opts<-list("algorithm"="NLOPT_LN_COBYLA",xtol_rel=1e-15)
options=list("algorithm"="NLOPT_LN_AUGLAG",xtol_rel=1e-15,maxeval=16000,"local_opts"=local_opts)
# R0=778.1259
# L=36
out_1a=nloptr(x0=35,
              eval_f=opt_fct,
              opts=options,
              lb=0,
              eval_g_eq = constraint,
              a=40,
              b=1,
              c=0,
              discount=0.05,
              int_state=778.1259)

out_1a$solution
-out_1a$objective
```


```{r}
table_function <- function(choice, a=40, b=1, c=0, discount=0.05, int_state=778.1259) {
  t=choice
  P=vector(mode="numeric", length=0) # P is price
  Q=vector(mode="numeric", length=0) # This is q
  R=vector(mode="numeric", length=0) # State is reserves
  benefits=vector(mode="numeric", length=0)
  lambda=vector(mode="numeric", length=0)

  # Define initial conditions. Since we are working down these are at the end of the time frame
  P[t+1]=a
  R[t+1]=0
  Q[t+1]=0
  benefits[t+1]=0

  for(i in (t+1):2) {

    P[i-1]=(P[i]-(c*discount))/(1+discount) # Gives P from t+1 to 1

    Q[i-1]=(a-P[i-1])/b # Gives Q from t+1 to 1

  }
  
  benefits[1]=(P[1]-c)*Q[1]
  lambda[1]=0
  R[1]=int_state
  rho=1/(1+discount)
  lambda[t+1]=(P[t]-c)/rho
  
  for(i in 2:((t+1)-1)) {
    R[i]=R[i-1]-Q[i-1] # Gives R from 2 to t+1

    benefits[i]=(P[i]-c)*Q[i] # Gives benefits from 2 to t+1
    lambda[i]=(P[i-1]-c)/(1/(1+discount))
  }

  time=seq(from=0,to=t)
  pv=rho^time*benefits
  npv=sum(pv)
  
  return(data.frame("Time" = time, "Price"=P, "Quantity"=Q, "Reserves" = round(R), "Profits" = benefits, "Lambda"=lambda, "PV"=pv)
    )
}

options(scipen=999)
df <- table_function(out_1a$solution)

knitr::kable(df, align="c")
quantity=ggplot(df, aes(x = -Quantity, y = Price)) +
  geom_point()
price=ggplot(df, aes(x = Time, y = Price)) +
  geom_point()
reserves=ggplot(df, aes(x = -Time,  y= -Reserves)) +
  geom_point() + coord_flip()
time=ggplot(df, aes(x = Time, y = -Time)) +
  geom_point()
cowplot::plot_grid(quantity, price, reserves, time, nrow=2,ncol=2)

```



<span style="color: blue;">
</span>

### B.

Is the resource getting more scarce over time? What is your evidence?

<span style="color: blue;">**The resource is getting more scarce over time. As reserves diminish, quantity extracted lessens, and the price goes up. This is  indicated by lambda, which for a given time period reflects the value of one more unit of the resource in the next time period. In this case, Lambda increases over time, reflecting the increasing scarcity of the resource.**</span>

<span style="color: blue;">
</span>

### C.

If you were to increase the initial stock of reserve ($R_0$) by one unit and resolve the equilibrium, how much would the present discounted value of profits increase?  Hint: you don't need to resolve the model. An approximate answer is already in your table.

<span style="color: blue;">**Solution**</span>

<span style="color: blue;">
</span>


## Question 2: Exhaustible resource extraction with costs {.tabset}

Consider three extensions of the model in question 1.  In case 1, there is a constant marginal cost of extraction c = 3.178025.  The profits in each year are now $\pi_t=(p_t-c)q_t$ and the shadow price satisfies $p_t-c=\rho\lambda_{t+1}$. In case 2, marginal costs are zero again but the initial stock of reserves increases to $R_0$ = 948.2253 units.  In case 3, marginal costs are still zero and reserves go back to $R_0$=778.1259, but the demand curve parameters are now a=40, b=0.75194.  A lower value of b means that the demand curve rotates outward around a=40.  

### A.

What is the final year T for the three cases? Explain why they are different from question 1.

<span style="color: blue;">**The final year T for each of the three cases are: case 1=34, case 2=40, and case 3=30.**</span>

```{r}
# Im not sure if these are correct... These solutions input into the table function give weird answers 
out_2a1=nloptr(x0=40,
               eval_f=opt_fct,
               opts=options,
               lb=0,
               eval_g_eq = constraint,
               a=40,
               b=1,
               c=3.178025,
               discount=0.05,
               int_state=778.1259)
out_2a1$solution
-out_2a1$objective

out_2a2=nloptr(x0=35,
               eval_f=opt_fct,
               opts=options,
               lb=0,
               eval_g_eq = constraint,
               a=40,
               b=1,
               c=0,
               discount=0.05,
               int_state=948.2253)
out_2a2$solution
-out_2a2$objective

out_2a3=nloptr(x0=25,
               eval_f=opt_fct,
               opts=options,
               lb=0,
               eval_g_eq = constraint,
               a=40,
               b=0.75194,
               c=0,
               discount=0.05,
               int_state=778.1259)
out_2a3$solution
-out_2a3$objective



```




<span style="color: blue;">
</span>

### B. 

How does $\lambda_1$ in the three cases compare to its value in question 1?  What explains the differences?

<span style="color: blue;">**Solution**</span>

```{r}
# This is correct for each case except for case 1. Not sure why it isn't depleting the resource to zero aside from it looks like it could use another year.
df2b1 <- table_function(out_2a1$solution, c=3.178025)
knitr::kable(df2b1, align='c')
df2b2 <- table_function(out_2a2$solution, int_state=948.2253)
knitr::kable(df2b2, align='c')
df2b3 <- table_function(out_2a3$solution, b=0.75194)
knitr::kable(df2b3, align='c')
```

<span style="color: blue;">
</span>


### C.

What happens to the present discounted value of profits (in other words, the value of the reserves $R_0$) in the three cases compared to question 1? Explain the differences.

<span style="color: blue;">**Solution**</span>

<span style="color: blue;">
</span>

